---
title: "DGE_GSEA"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(readxl)
library(biomaRt)
library(data.table)
library(dplyr)
library(ggrepel)
library(ggplot2)
library(readr)
library(stringr)
library(tidyr)
library(devtools)
library(DESeq2)
library(tidyverse)
library(edgeR)
library(writexl)
library(clusterProfiler)
library(org.Hs.eg.db)
library(svglite)

knitr::opts_knit$set(root.dir = "/Users/cindywang/Desktop/Kramer/Publication/Data")
```

# Sex

```{r}
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
sex<-read.csv("sex(autosomal)_ENSEMBL(in).csv")

# add pre-count sig genes on top of plot 
sex$case<-as.factor(sex$case)
count_sex<-sex %>% dplyr::count(case)
annot_df<-count_sex %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = max(-log10(sex$p_value)) * 1.05
  )

##p value 
ggplot(data = sex, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 1.2,alpha=1) +
  scale_colour_manual(
    values = c(
      "A" = "#92c5de",
      "B" = "#0571b0",
      "C"="#fdae61",
      "D"="#ca0020",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "101 Female vs 148 Male"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5))+
  xlim(-1,1)+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )
# ggsave("Combined_DREAM_sex_pub.png",width=6,height=6,unit="in",dpi = 300)

ggsave(
  "Combined_DREAM_sex_pub.svg",
  device = svglite,
  width = 6,
  height = 6
)
```

## Find DEG 

```{r}
# isolate the interesting columns 
sex_deg<-sex %>%
  filter(case %in% c("A","B","C","D")) %>%
  dplyr::select(geneid,log_fc,adj_p_val,case)

# edit 
sex_deg<-sex_deg %>%
  mutate(
    LFC=round(log_fc,2),
    Adjusted_p_value = case_when(
      adj_p_val<0.0005~"<0.0005",
      adj_p_val<0.005~"<0.005",
      adj_p_val<0.05~"<0.05",
      TRUE~as.character(adj_p_val)
    ),
    Category=recode(
      case,
      "A"="Moderately upregulated in male",
      "B"="Strongly upregulated in male",
      "C"="Moderately upregulated in female",
      "D"="Strongly upregulated in female"
    )) %>%
  dplyr::select(-log_fc,-adj_p_val,-case) %>%
  arrange(desc(LFC))


# convert ENSEMBL ID to gene symbol
mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl"
)

# Query Ensembl
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = sex_deg$geneid,
  mart = mart
)


#combine deg and gene_map based on ensembl ID 
sex_deg_gene_name<- sex_deg %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

write_xlsx(sex_deg_gene_name, "sex_deg.xlsx")
```

## GSEA
leading edge genes

```{r}
res<-sex

ranks <- res$t
names(ranks) <- res$geneid
ranks <- sort(ranks, decreasing = TRUE)

set.seed(42)  # gseGO uses permutation-based fgsea internally; fix seed for reproducibility
gobp_gsea_sex <- gseGO(
  geneList      = ranks,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  keyType       = "ENSEMBL",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  eps           = 0,
  seed          = TRUE,
  verbose       = FALSE
)

#plotting 
df <- as.data.frame(gobp_gsea_sex)

n_show <- 4
plot_df <- df %>%
  arrange(desc(NES)) %>%
  slice_head(n = n_show) %>%
  bind_rows(df %>% arrange(NES) %>% slice_head(n = n_show)) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  mutate(fill_group = case_when(
    NES > 0 & p.adjust <= 0.05 ~ "Up_sig",
    NES > 0 & p.adjust > 0.05 ~ "Up_ns",
    NES < 0 & p.adjust <= 0.05 ~ "Down_sig",
    TRUE ~ "Down_ns"
  )) 


sex_gsea<-ggplot(plot_df, aes(x = NES, y = Description, fill = fill_group)) +
  geom_col(width = 0.85) +
  scale_fill_manual(values = c(
    "Up_sig"   = "#2c7bb6",
    "Up_ns"    = "#abd9e9",
    "Down_sig" = "#d7191c",
    "Down_ns"  = "#fdae61"
  )) +
  geom_vline(xintercept = 0) +
  labs(x = "Normalized Enrichment Score", y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank()
  ) +
  scale_x_continuous(
    limits = c(-3, 3),
    breaks = c(-2, 0, 2)
  )

# Save the exact same plot
ggsave(
  filename = "sex_GSEA.pdf",
  plot = sex_gsea,
  width = 12,
  height = 6,
  units = "in"
)

```

## leading edge genes

```{r}
#### thoracic lumbar #### 
gsea_res <- as.data.frame(gobp_gsea_sex)

# Top 4 positive and top 4 negative pathways
top_pathways <- bind_rows(
  gsea_res %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 4),

  gsea_res %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 7)
)

# Create a 2-column data frame
leading_edge_df <- top_pathways %>%
  rowwise() %>%
  mutate(
    Leading_edge_genes = {
      ensembl <- unlist(strsplit(core_enrichment, "/"))
      symbols <- mapIds(
        org.Hs.eg.db,
        keys = ensembl,
        keytype = "ENSEMBL",
        column = "SYMBOL",
        multiVals = "first"
      )
      paste(na.omit(symbols), collapse = ", ")
    }
  ) %>%
  ungroup() %>%
  dplyr::select(
    Pathway = Description,
    Leading_edge_genes
  )

View(leading_edge_df)
# write.csv(leading_edge_df, "leg_sex.csv", row.names = FALSE)
```

## validation cohort 

include sex chromosome genes 
```{r}
#find which sex-chr genes are differentially expressed 
library(biomaRt)
mart<-useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  mirror = "useast"
)

sex_chr_df<-getBM(
  attributes = c(
    "ensembl_gene_id",
    "hgnc_symbol",
    "chromosome_name"
  ),
  filters="chromosome_name",
  values=c("X","Y"),
  mart = mart
)

sex_chr_df<-subset(sex_chr_df,hgnc_symbol!="")
sex_genes<-unique(sex_chr_df$hgnc_symbol)

sex_chr_genes_df<-data.frame(
  gene_name=sex_genes,
  stringsAsFactors = FALSE
)

# load data 
sex_val<-read.csv("validtion(include_sex_chr_genes).csv")

# #add pre-count sig genes on top of plot 
sex_val$case<-as.factor(sex_val$case)
count_sex<-sex_val %>% dplyr::count(case)
annot_df<-count_sex %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = max(-log10(sex_val$p_value)) * 1.05
  )

# pull sig genes 
sig_genes<-sex_val %>%
  dplyr::filter(case %in% c("A","B","C","D")) %>%
  dplyr::pull(geneid)

# find overlap with sex chro genes 
overlap_sex_chr<-intersect(sig_genes,sex_chr_genes_df$gene_name)
length(overlap_sex_chr) #50 

#label the significant, protein-coding genes 
highlight_df <- sex_val %>% filter(geneid %in% overlap_sex_chr)

##p value 
ggplot(data = sex_val, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 1.2,alpha=1) +
  scale_colour_manual(
    values = c(
      "A" = "#92c5de",
      "B" = "#0571b0",
      "C"="#fdae61",
      "D"="#ca0020",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "101 Female vs 148 Male (including sex chromosome genes)"
  )+  
  geom_text_repel(
    data = highlight_df,
    aes(label = geneid),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5))+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )

# ggsave("validation_sex_chr_genes.png",width=10,height=8,unit="in",dpi = 300)
```

#Level

```{r}
cerv_hox<-c("HOXC4",
           "HOXA5","HOXB5","HOXC5",
           "HOXA6","HOXB6","HOXC6",
           "HOXA7","HOXB7",
           "HOXB8","HOXC8","HOXD8")

thor_hox<-c("HOXC8","HOXD8",
           "HOXA9","HOXC9","HOXD9")

lum_hox<-c("HOXD9",
           "HOXA10", "HOXC10", "HOXD10",
           "HOXA11","HOXC11", "HOXD11")

# convert ENSEMBL ID to gene symbol
mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl"
)

# Combine all HOX genes and remove duplicates
hox_genes <- unique(c(cerv_hox, thor_hox, lum_hox))

# Query Ensembl
gene_map <- getBM(
  attributes = c("hgnc_symbol","ensembl_gene_id"),
  filters = "hgnc_symbol",
  values = hox_genes,
  mart = mart
)

#match the corresponding ENSEMBL IDs for each HOX gene
cerv_hox_ensembl<-gene_map %>%
  filter(hgnc_symbol %in% cerv_hox) %>%
  pull(ensembl_gene_id)

thor_hox_ensembl<-gene_map %>%
  filter(hgnc_symbol %in% thor_hox) %>%
  pull(ensembl_gene_id)

lum_hox_ensembl<-gene_map %>%
  filter(hgnc_symbol %in% lum_hox) %>%
  pull(ensembl_gene_id)

```


```{r}
####### lumbar vs cervical #######
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
lumcerv<-read.csv("cerv_lumb(refCerv)_ENSEMBL(in).csv") %>%
  as.data.frame()

#add pre-count sig genes on top of plot 
lumcerv$case<-as.factor(lumcerv$case)
count_lumcerv<-lumcerv %>% dplyr::count(case)
annot_df<-count_lumcerv %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = max(-log10(lumcerv$p_value)) * 1.1
  )

#highlight 
genes_to_label <- c(cerv_hox_ensembl,lum_hox_ensembl) # HOX genes
highlight_df <- lumcerv %>%
  filter(geneid %in% genes_to_label) %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

#p value
ggplot(data = lumcerv, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
     "A" = "#5F9FA0",
      "B" = "#003F3F",
      "C"="#F2A463",
      "D"="#FF6f61",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "195 Cervical vs 43 Lumbar"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  geom_text_repel(
    data = highlight_df,
    aes(label = hgnc_symbol),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  xlim(-4,4.5)+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )

# ggsave(
#   "DREAM_cervlum_pub.pdf",
#   width = 6,
#   height = 6,
# )

####### Cervical vs thoracic #######
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
cervthor<-read.csv("cerv_thor(refCerv)_ENSEMBL(in).csv") %>%
  as.data.frame()

sum(cervthor$case =="A" | cervthor$case =="B") 
sum(cervthor$case =="C" | cervthor$case =="D") 

#add pre-count sig genes on top of plot 
cervthor$case<-as.factor(cervthor$case)
count_cervthor<-cervthor %>% dplyr::count(case)
annot_df<-count_cervthor %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = 15
  )

genes_to_label <- c(cerv_hox_ensembl,thor_hox_ensembl)
highlight_df <- cervthor %>% 
  filter(geneid %in% genes_to_label) %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

##p value 
ggplot(data = cervthor, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
      "A" = "#5F9FA0",
      "B" = "#003F3F",
      "C"="#F2A463",
      "D"="#FF6f61",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "195 Cervical vs 11 Thoracic"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  xlim(-4,4.5)+
  geom_text_repel(
    data = highlight_df,
    aes(label = hgnc_symbol),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )
# ggsave(
#   "DREAM_cervthora.pdf",
#   width = 6,
#   height = 6
# )

######## thoracic & lumbar
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
thorlum<-read.csv("thor_lum(refThor)_ENSEMBL(in).csv")%>%
  as.data.frame()

sum(thorlum$case =="A" | thorlum$case =="B") 
sum(thorlum$case =="C" | thorlum$case =="D") 

#add pre-count sig genes on top of plot 
thorlum$case<-as.factor(thorlum$case)
count_thorlum<-thorlum %>% dplyr::count(case)
annot_df<-count_thorlum %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = 15
  )

# highlight 
genes_to_label <- c(lum_hox_ensembl,thor_hox_ensembl) # HOX genes
highlight_df <- thorlum %>% 
  filter(geneid %in% genes_to_label) %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

#plot 
ggplot(data = thorlum, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
      "A" = "#5F9FA0",
      "B" = "#003F3F",
      "C"="#F2A463",
      "D"="#FF6f61",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "11 Thoracic vs 43 Lumbar"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  xlim(-4,4.5)+
  geom_text_repel(
    data = highlight_df,
    aes(label = hgnc_symbol),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  ylim(0,15)+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )

# ggsave(
#   "thorlum_test.pdf",
#   width = 6,
#   height = 6
# )

```

## Find DEG

```{r}
###### cervical thoracic ######
cervthor_deg<-cervthor %>%
  filter(case %in% c("A","B","C","D")) %>%
  dplyr::select(geneid,log_fc,adj_p_val,case)

cervthor_deg<-cervthor_deg %>%
  mutate(
    LFC=round(log_fc,2),
    Adjusted_p_value = case_when(
      adj_p_val<0.0005~"<0.0005",
      adj_p_val<0.005~"<0.005",
      adj_p_val<0.05~"<0.05",
      TRUE~as.character(adj_p_val)
    ),
    Category=recode(
      case,
      "A"="Moderately upregulated in thoracic",
      "B"="Strongly upregulated in thoracic",
      "C"="Moderately upregulated in cervical",
      "D"="Strongly upregulated in cervical"
    )) %>%
  dplyr::select(-log_fc,-adj_p_val,-case) %>%
  arrange(desc(LFC))

# edit 
# convert ENSEMBL ID to gene symbol
# mart <- useEnsembl(
#   biomart = "genes",
#   dataset = "hsapiens_gene_ensembl"
# )

# Query Ensembl
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = cervthor_deg$geneid,
  mart = mart
)

#combine deg and gene_map based on ensembl ID 
cervthor_deg_gene_name<- cervthor_deg %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )
write_xlsx(cervthor_deg_gene_name, "cervthor_deg.xlsx")




###### lumbar cervical ######
lumcerv_deg<-lumcerv %>%
  filter(case %in% c("A","B","C","D")) %>%
  dplyr::select(geneid,log_fc,adj_p_val,case)

lumcerv_deg<-lumcerv_deg %>%
  mutate(
    LFC=round(log_fc,2),
    Adjusted_p_value = case_when(
      adj_p_val<0.0005~"<0.0005",
      adj_p_val<0.005~"<0.005",
      adj_p_val<0.05~"<0.05",
      TRUE~as.character(adj_p_val)
    ),
    Category=recode(
      case,
      "A"="Moderately upregulated in lumbar",
      "B"="Strongly upregulated in lumbar",
      "C"="Moderately upregulated in cervical",
      "D"="Strongly upregulated in cervical"
    )) %>%
  dplyr::select(-log_fc,-adj_p_val,-case) %>%
  arrange(desc(LFC))%>%
  dplyr::rename(Gene=geneid)

# Query Ensembl
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = lumcerv_deg$Gene,
  mart = mart
)

#combine deg and gene_map based on ensembl ID 
lumcerv_deg_gene_name<- lumcerv_deg %>%
  left_join(
    gene_map,
    by = c("Gene" = "ensembl_gene_id")
  )
write_xlsx(lumcerv_deg_gene_name, "Lumbcerv_deg.xlsx")




###### thoracic lumbar ######
thorlum_deg<-thorlum %>%
  filter(case %in% c("A","B","C","D")) %>%
  dplyr::select(geneid,log_fc,adj_p_val,case)

thorlum_deg<-thorlum_deg %>%
  mutate(
    LFC=round(log_fc,2),
    Adjusted_p_value = case_when(
      adj_p_val<0.0005~"<0.0005",
      adj_p_val<0.005~"<0.005",
      adj_p_val<0.05~"<0.05",
      TRUE~as.character(adj_p_val)
    ),
    Category=recode(
      case,
      "A"="Moderately upregulated in lumbar",
      "B"="Strongly upregulated in lumbar",
      "C"="Moderately upregulated in thoracic",
      "D"="Strongly upregulated in thoracic"
    )) %>%
  dplyr::select(-log_fc,-adj_p_val,-case) %>%
  arrange(desc(LFC))%>%
  dplyr::rename(Gene=geneid)

# Query Ensembl
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = thorlum_deg$Gene,
  mart = mart
)

#combine deg and gene_map based on ensembl ID 
thorlum_deg_gene_name<- thorlum_deg %>%
  left_join(
    gene_map,
    by = c("Gene" = "ensembl_gene_id")
  )

write_xlsx(thorlum_deg_gene_name, "thorlum_deg.xlsx")
```

## GSEA Level 

```{r}
## lumbar cervical comparison 
res<-lumcerv

# use t stats as ranking metric 
ranks <- res$t
names(ranks) <- res$geneid
ranks <- sort(ranks, decreasing = TRUE)

set.seed(1)  
gobp_gsea_cervlum <- gseGO(
  geneList      = ranks,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  keyType       = "ENSEMBL",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  eps           = 0,
  seed          = TRUE,
  verbose       = FALSE
)

#plotting 
df <- as.data.frame(gobp_gsea_cervlum)

n_show <- 4
plot_df <- df %>%
  arrange(desc(NES)) %>%
  slice_head(n = n_show) %>%
  bind_rows(df %>% arrange(NES) %>% slice_head(n = n_show)) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  mutate(fill_group = case_when(
    NES > 0 & p.adjust <= 0.05 ~ "Up_sig",
    NES > 0 & p.adjust > 0.05 ~ "Up_ns",
    NES < 0 & p.adjust <= 0.05 ~ "Down_sig",
    TRUE ~ "Down_ns"
  )) 

lumcerv<-ggplot(plot_df, aes(x = NES, y = Description, fill = fill_group)) +
  geom_col(width = 0.85) +
  scale_fill_manual(values = c(
  "Up_sig"   = "#FF8822",
  "Up_ns"    = "#abd9e9",
  "Down_sig" = "#ffc30b",
  "Down_ns"  = "#008000"
  ))+
  geom_vline(xintercept = 0) +
  labs(x = "Normalized Enrichment Score", y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.major.y = element_blank())+ 
  scale_x_continuous(breaks = c(-2.5,0, 2.5)) + 
  xlim(-4.5,4.5)

ggsave(
  filename = "lumcerv_GSEA.pdf",
  plot = lumcerv,
  width = 12,
  height = 6,
  units = "in"
)




## cervical thoracic comparison 
res<- cervthor

# use t stats as ranking metric
ranks <- res$t
names(ranks) <- res$geneid
ranks <- sort(ranks, decreasing = TRUE)

set.seed(40)
gobp_gsea_cervthor <- gseGO(
  geneList      = ranks,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  keyType       = "ENSEMBL",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  eps           = 0,
  seed          = TRUE,
  verbose       = FALSE
)

#plotting 
df <- as.data.frame(gobp_gsea_cervthor)

n_show <- 4
plot_df <- df %>%
  arrange(desc(NES)) %>%
  slice_head(n = n_show) %>%
  bind_rows(df %>% arrange(NES) %>% slice_head(n = n_show)) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  mutate(fill_group = case_when(
    NES > 0 & p.adjust <= 0.05 ~ "Up_sig",
    NES > 0 & p.adjust > 0.05 ~ "Up_ns",
    NES < 0 & p.adjust <= 0.05 ~ "Down_sig",
    TRUE ~ "Down_ns"
  )) 

pdf("cervthor_GSEA.pdf", width=12, height=6)

ggplot(plot_df, aes(x = NES, y = Description, fill = fill_group)) +
  geom_col(width = 0.85) +
  scale_fill_manual(values = c(
  "Up_sig"   = "#FF8822",
  "Up_ns"    = "#abd9e9",
  "Down_sig" = "#ffc30b",
  "Down_ns"  = "#008000"
  ))+
  geom_vline(xintercept = 0) +
  labs(x = "Normalized Enrichment Score", y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", 
        panel.grid.major.y = element_blank())+ 
  xlim(-4,4.2)

dev.off()




######### thoracic lumbar ######### 
res<- thorlum

# use t stats as ranking metric
ranks <- res$t
names(ranks) <- res$geneid
ranks <- sort(ranks, decreasing = TRUE)

set.seed(40)
gobp_gsea_thorlum <- gseGO(
  geneList      = ranks,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  keyType       = "ENSEMBL",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  eps           = 0,
  seed          = TRUE,
  verbose       = FALSE
)

#plotting 
df <- as.data.frame(gobp_gsea_thorlum)

n_show <- 4
plot_df <- df %>%
  arrange(desc(NES)) %>%
  slice_head(n = n_show) %>%
  bind_rows(df %>% arrange(NES) %>% slice_head(n = n_show)) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  mutate(fill_group = case_when(
    NES > 0 & p.adjust <= 0.05 ~ "Up_sig",
    NES > 0 & p.adjust > 0.05 ~ "Up_ns",
    NES < 0 & p.adjust <= 0.05 ~ "Down_sig",
    TRUE ~ "Down_ns"
  )) 

pdf("thorlum_GSEA.pdf", width=12, height=6)

ggplot(plot_df, aes(x = NES, y = Description, fill = fill_group)) +
  geom_col(width = 0.85) +
  scale_fill_manual(values = c(
  "Up_sig"   = "#FF8822",
  "Up_ns"    = "#abd9e9",
  "Down_sig" = "#ffc30b",
  "Down_ns"  = "#008000"
  ))+
  geom_vline(xintercept = 0) +
  labs(x = "Normalized Enrichment Score", y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", 
        panel.grid.major.y = element_blank())+ 
  xlim(-4.5,4.5)

dev.off()

```
## leading edge genes

```{r}
#### thoracic lumbar #### 
gsea_res <- as.data.frame(gobp_gsea_thorlum)

# Top 4 positive and top 4 negative pathways
top_pathways <- bind_rows(
  gsea_res %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 4),

  gsea_res %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 4)
)

# Create a 2-column data frame
leading_edge_df <- top_pathways %>%
  rowwise() %>%
  mutate(
    Leading_edge_genes = {
      ensembl <- unlist(strsplit(core_enrichment, "/"))
      symbols <- mapIds(
        org.Hs.eg.db,
        keys = ensembl,
        keytype = "ENSEMBL",
        column = "SYMBOL",
        multiVals = "first"
      )
      paste(na.omit(symbols), collapse = ", ")
    }
  ) %>%
  ungroup() %>%
  dplyr::select(
    Pathway = Description,
    Leading_edge_genes
  )

leading_edge_df
write.csv(leading_edge_df, "leg_thorlum.csv", row.names = FALSE)

getwd()

#### cervical thoracic #### 
gsea_res <- as.data.frame(gobp_gsea_cervthor)

# Top 4 positive and top 4 negative pathways
top_pathways <- bind_rows(
  gsea_res %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 4),

  gsea_res %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 4)
)

# Create a 2-column data frame
leading_edge_df <- top_pathways %>%
  rowwise() %>%
  mutate(
    Leading_edge_genes = {
      ensembl <- unlist(strsplit(core_enrichment, "/"))
      symbols <- mapIds(
        org.Hs.eg.db,
        keys = ensembl,
        keytype = "ENSEMBL",
        column = "SYMBOL",
        multiVals = "first"
      )
      paste(na.omit(symbols), collapse = ", ")
    }
  ) %>%
  ungroup() %>%
  dplyr::select(
    Pathway = Description,
    Leading_edge_genes
  )

leading_edge_df
write.csv(leading_edge_df, "leg_cervthor.csv", row.names = FALSE)



#### cervical lumbar ####
gsea_res <- as.data.frame(gobp_gsea_cervlum)

# Top 4 positive and top 4 negative pathways
top_pathways <- bind_rows(
  gsea_res %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 4),

  gsea_res %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 4)
)

# Create a 2-column data frame
leading_edge_df <- top_pathways %>%
  rowwise() %>%
  mutate(
    Leading_edge_genes = {
      ensembl <- unlist(strsplit(core_enrichment, "/"))
      symbols <- mapIds(
        org.Hs.eg.db,
        keys = ensembl,
        keytype = "ENSEMBL",
        column = "SYMBOL",
        multiVals = "first"
      )
      paste(na.omit(symbols), collapse = ", ")
    }
  ) %>%
  ungroup() %>%
  dplyr::select(
    Pathway = Description,
    Leading_edge_genes
  )

leading_edge_df
write.csv(leading_edge_df, "leg_cervlum.csv", row.names = FALSE)


```

#Age (continuous)

```{r}
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
age<-read.csv("age_cont_ENSEMBL(in).csv")

# age$case<-as.factor(age$case)
count_age<-age %>% dplyr::count(case)
annot_df<-count_age %>%
  mutate(
    x=case_when(
      case=="A"~0.02,
      case=="B"~1.5,
      case=="C"~-0.02,
      case=="D"~-1.5),
    y = max(-log10(age$p_value)) * 1.1
    )

#find the number of significant genes with age
sum(age$adj_p_val < 0.05 & age$case =="A") # 0 significant genes with age as a continuous variable
sum(age$adj_p_val < 0.05 & age$case =="C")

##p value 
ggplot(data = age, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
      "A" = "#92c5de",
      "B" = "blue",
      "C"="#fdae61",
      "D"="#ca0020",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "Age"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )
# ggsave("Combined_DREAM_age_highlight.png",width=6,height=6,unit="in",dpi = 300)

ggsave(
  "Combined_DREAM_age_pub.svg",
  device = svglite,
  width = 6,
  height = 6
)
```

## Find DEG 

```{r}
# isolate the interesting columns
age_deg<-age %>%
  filter(case %in% c("A","B","C","D")) %>%
  dplyr::select(geneid,log_fc,adj_p_val,case)

# edit 
age_deg<-age_deg %>%
  mutate(
    LFC=round(log_fc,2),
    Adjusted_p_value = case_when(
      adj_p_val<0.0005~"<0.0005",
      adj_p_val<0.005~"<0.005",
      adj_p_val<0.05~"<0.05",
      TRUE~as.character(adj_p_val)
    ),
    Category=recode(
      case,
      "A"="Moderately upregulated in older individuals",
      "B"="Strongly upregulated in older individuals",
      "C"="Moderately upregulated in younger individuals",
      "D"="Strongly upregulated in younger individuals"
    )) %>%
  dplyr::select(-log_fc,-adj_p_val,-case) %>%
  arrange(desc(LFC)) 

# convert ENSEMBL ID to gene symbol
# mart <- useEnsembl(
#   biomart = "genes",
#   dataset = "hsapiens_gene_ensembl"
# )

# Query Ensembl
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = age_deg$geneid,
  mart = mart
)

#combine deg and gene_map based on ensembl ID 
age_deg_gene_name<- age_deg %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )
# write_xlsx(age_deg_gene_name, "age_deg.xlsx")
```

## GSEA age

```{r}
res<-age

# Prefer the model's test statistic for ranking: t-statistic
ranks <- res$t
names(ranks) <- res$geneid
ranks <- sort(ranks, decreasing = TRUE)

# Sanity checks before running
sum(duplicated(ranks))        # 1 tied values?
sum(is.infinite(ranks))       # any Inf from p=0?
length(ranks)                 # how many genes going in?

set.seed(40)  
gobp_gsea_age <- gseGO(
  geneList      = ranks,
  OrgDb         = org.Hs.eg.db,
  ont           = "BP",
  keyType       = "ENSEMBL",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 1,
  pAdjustMethod = "BH",
  eps           = 0,
  seed          = TRUE,
  verbose       = FALSE
)

#plotting 
df <- as.data.frame(gobp_gsea_age)

n_show <- 4
plot_df <- df %>%
  arrange(desc(NES)) %>%
  slice_head(n = n_show) %>%
  bind_rows(df %>% arrange(NES) %>% slice_head(n = n_show)) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  mutate(fill_group = case_when(
    NES > 0 & p.adjust <= 0.05 ~ "Up_sig",
    NES > 0 & p.adjust > 0.05 ~ "Up_ns",
    NES < 0 & p.adjust <= 0.05 ~ "Down_sig",
    TRUE ~ "Down_ns"
  )) 


pdf("age_GSEA_pub.pdf", width=12, height=6)

ggplot(plot_df, aes(x = NES, y = Description, fill = fill_group)) +
  geom_col(width = 0.85) +
  scale_fill_manual(values = c(
  "Up_sig"   = "#E4ADA9",
  "Up_ns"    = "#abd9e9",
  "Down_sig" = "#8EA604",
  "Down_ns"  = "#fdae61"
  ))+
  geom_vline(xintercept = 0) +
  labs(x = "Normalized Enrichment Score", y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.major.y = element_blank())

dev.off()

```
## leading edge genes

```{r}
gsea_res <- as.data.frame(gobp_gsea_age)

# Top 4 positive and top 4 negative pathways
top_pathways <- bind_rows(
  gsea_res %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 4),

  gsea_res %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 4)
)

# Create a 2-column data frame
leading_edge_df <- top_pathways %>%
  rowwise() %>%
  mutate(
    Leading_edge_genes = {
      ensembl <- unlist(strsplit(core_enrichment, "/"))
      symbols <- mapIds(
        org.Hs.eg.db,
        keys = ensembl,
        keytype = "ENSEMBL",
        column = "SYMBOL",
        multiVals = "first"
      )
      paste(na.omit(symbols), collapse = ", ")
    }
  ) %>%
  ungroup() %>%
  dplyr::select(
    Pathway = Description,
    Leading_edge_genes
  )

View(leading_edge_df)

# write.csv(leading_edge_df, "leg_age.csv", row.names = FALSE)
```
# Sensitivity Analysis 

```{r}
####### lumbar vs cervical #######
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
sens_lumcerv<-read.csv("sens_cervlum(refCerv).csv") %>%
  as.data.frame()

# #add pre-count sig genes on top of plot 
sens_lumcerv$case<-as.factor(sens_lumcerv$case)
count_lumcerv<-sens_lumcerv %>% dplyr::count(case)
annot_df<-count_lumcerv %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = max(-log10(sens_lumcerv$p_value)) * 1.1
  )

#highlight 
genes_to_label <- c(cerv_hox_ensembl,lum_hox_ensembl) # HOX genes
highlight_df <- sens_lumcerv %>%
  filter(geneid %in% genes_to_label) %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

##p value
ggplot(data = sens_lumcerv, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
     "A" = "#92c5de",
      "B" = "blue",
      "C"="#fdae61",
      "D"="#ca0020",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "36 Cervical vs 33 Lumbar"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  geom_text_repel(
    data = highlight_df,
    aes(label = hgnc_symbol),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  xlim(-4,4.5)+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )
# ggsave("sens_cervlum_pub.png",width=6,height=6,unit="in",dpi = 300)


####### Cervical vs thoracic #######
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
sens_cervthor<-read.csv("sens_cervthor(refCerv).csv") %>%
  as.data.frame()

#add pre-count sig genes on top of plot 
sens_cervthor$case<-as.factor(sens_cervthor$case)
count_cervthor<-sens_cervthor %>% dplyr::count(case)
annot_df<-count_cervthor %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = 15
  )

genes_to_label <- c(cerv_hox_ensembl,thor_hox_ensembl)
highlight_df <- sens_cervthor %>% 
  filter(geneid %in% genes_to_label) %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

##p value 
ggplot(data = sens_cervthor, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
      "A" = "#92c5de",
      "B" = "blue",
      "C"="#fdae61",
      "D"="#ca0020",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "36 Cervical vs 11 Thoracic"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  xlim(-4,4.5)+
  geom_text_repel(
    data = highlight_df,
    aes(label = hgnc_symbol),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )
# ggsave("sens_cervthora.png",width=6,height=6,unit="in",dpi = 300)


######## thoracic & lumbar
setwd("/Users/cindywang/Desktop/Kramer/Publication/Data")
sens_thorlum<-read.csv("sens_thorlum(refThor).csv")%>%
  as.data.frame()

#add pre-count sig genes on top of plot 
sens_thorlum$case<-as.factor(sens_thorlum$case)
count_thorlum<-sens_thorlum %>% dplyr::count(case)
annot_df<-count_thorlum %>%
  mutate(
    x=case_when(
      case=="A"~0.5,
      case=="B"~1.5,
      case=="C"~-0.5,
      case=="D"~-1.5),
    y = 15
  )

# highlight 
genes_to_label <- c(lum_hox_ensembl,thor_hox_ensembl) # HOX genes
highlight_df <- sens_thorlum %>% 
  filter(geneid %in% genes_to_label) %>%
  left_join(
    gene_map,
    by = c("geneid" = "ensembl_gene_id")
  )

#plot 
ggplot(data = sens_thorlum, aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(color = case), size = 0.8) +
  scale_colour_manual(
    values = c(
      "A" = "#92c5de",
      "B" = "blue",
      "C"="#fdae61",
      "D"="#ca0020",
      "E" = "gray"
    )
  ) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(
    y = expression(-log[10]~P),
    x = expression(log[2]~fold~change),
    subtitle = "11 Thoracic vs 33 Lumbar"
  )+
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        legend.text  = element_text(size = 20),
        title = element_text(size=20),
        plot.subtitle = element_text(size=20,hjust = 0.5)
        )+
  xlim(-4,4.5)+
  geom_text_repel(
    data = highlight_df,
    aes(label = hgnc_symbol),
    size = 5,
    fontface = "bold",
    color = "black",
    segment.color = "black",
    segment.size = 0.8,
    min.segment.length = 0,
    box.padding = 0.4,
    point.padding = 0.3
  )+
  ylim(0,15)+
  geom_text(
    data = annot_df,
    aes(x = x, y = y, label = n, color = case),
    size = 6,
    fontface = "bold",
    show.legend = FALSE
  )
# ggsave("sens_thorlum.png",width=6,height=6,unit="in",dpi = 300)
 
```
# Concordance analysis 
1. find DEGs from main analysis
2. match those DEGs to sens analysis (n=?)
3. find correlation, t, df, p value
4. color by concordance

```{r}
###########################################
############ cervival thoracic ############
###########################################

# Keep overlapping genes and relevant columns
cervthor_df <- cervthor %>%
  dplyr::select(geneid, log_fc_main = log_fc, case_main = case) %>%
  inner_join(
    sens_cervthor%>%
      dplyr::select(geneid, log_fc_sens = log_fc, case_sens = case),
    by = "geneid"
  ) %>%
  mutate(
    concordance = case_when(
      case_sens %in% c("A", "B") &
        case_main %in% c("A", "B")        ~ "Upregulated",
      case_sens %in% c("C", "D") &
        case_main %in% c("C", "D")        ~ "Downregulated",
      case_sens == "E" | case_main == "E" ~ "NS",
      TRUE ~ "Discordant"
    )
  )

# Pearson correlation test
cor_result <- cor.test(
  cervthor_df$log_fc_sens,
  cervthor_df$log_fc_main,
  method = "pearson"
)

# Extract statistics
cor_stats <- data.frame(
  n = nrow(cervthor_df),
  correlation = unname(cor_result$estimate),
  t_value = unname(cor_result$statistic),
  df = unname(cor_result$parameter),
  p_value = cor_result$p.value
)
cor_stats

# plot and color by concordance 
annotation <- sprintf(
  "Pearson r = %.3f",
  cor_stats$correlation
)

ggplot(
  cervthor_df,
  aes(
    x = log_fc_main,
    y = log_fc_sens,
    color = concordance
  )
) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_point(
    data = cervthor_df %>%
      filter(!concordance %in% c("Upregulated", "Downregulated")),
    color = "grey70",
    size = 2,
    alpha = 0.6
  ) +
  geom_point(
    data = cervthor_df %>%
      filter(concordance %in% c("Upregulated", "Downregulated")),
    aes(color = concordance),
    size = 2,
    alpha = 0.9
  ) +
  annotate(
    "text",
    x = -Inf, y = Inf,
    label = annotation,
    hjust = -0.05, vjust = 1.1,
    size = 4
  ) +
  scale_color_manual(
    values = c(
      "Upregulated"   = "#D73027",
      "Downregulated"   = "#2166AC",
      "Discordant" ="darkgreen",
      "NS" = "grey65"),
  breaks = c("Upregulated", "Downregulated","Discordant")
  ) +
  labs(
  x = expression(log[2]~"[fold change, main analysis]"),
  y = expression(log[2]~"[fold change, sensitivity analysis]"),
  subtitle = "Cervical vs. Thoracic",
  color = "Concordance"
  ) +  
  ylim(-2,2.5)+
  theme_classic(base_size = 13)
ggsave("concordance_cervthor.png",width=8,height=6,unit="in",dpi = 300)

###########################################
############ cervival lumbar ############
###########################################
# Keep overlapping genes and relevant columns
lumcerv_df <- lumcerv %>%
  dplyr::select(geneid, log_fc_main = log_fc, case_main = case) %>%
  inner_join(
    sens_lumcerv%>%
      dplyr::select(geneid, log_fc_sens = log_fc, case_sens = case),
    by = "geneid"
  ) %>%
  mutate(
    concordance = case_when(
      case_sens %in% c("A", "B") &
        case_main %in% c("A", "B")        ~ "Upregulated",
      case_sens %in% c("C", "D") &
        case_main %in% c("C", "D")        ~ "Downregulated",
      case_sens == "E" | case_main == "E" ~ "NS",
      TRUE ~ "Discordant"
    )
  )

# Pearson correlation test
cor_result <- cor.test(
  lumcerv_df$log_fc_sens,
  lumcerv_df$log_fc_main,
  method = "pearson"
)

# Extract statistics
cor_stats <- data.frame(
  n = nrow(lumcerv_df),
  correlation = unname(cor_result$estimate),
  t_value = unname(cor_result$statistic),
  df = unname(cor_result$parameter),
  p_value = cor_result$p.value
)
cor_stats

# plot and color by concordance 
annotation <- sprintf(
  "Pearson r = %.3f",
  cor_stats$correlation
)

ggplot(
  lumcerv_df,
  aes(
    x = log_fc_main,
    y = log_fc_sens,
    color = concordance
  )
) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_point(
    data = lumcerv_df %>%
      filter(!concordance %in% c("Upregulated", "Downregulated")),
    color = "grey70",
    size = 2,
    alpha = 0.6
  ) +
  geom_point(
    data = lumcerv_df %>%
      filter(concordance %in% c("Upregulated", "Downregulated")),
    aes(color = concordance),
    size = 2,
    alpha = 0.9
  ) +
  annotate(
    "text",
    x = -Inf, y = Inf,
    label = annotation,
    hjust = -0.05, vjust = 1.1,
    size = 4
  ) +
  scale_color_manual(
    values = c(
      "Upregulated"   = "#D73027",
      "Downregulated"   = "#2166AC",
      "Discordant" ="darkgreen",
      "NS" = "grey65"),
  breaks = c("Upregulated", "Downregulated","Discordant")
  ) +
  labs(
  x = expression(log[2]~"[fold change, main analysis]"),
  y = expression(log[2]~"[fold change, sensitivity analysis]"),
  subtitle = "Cervical vs. Lumbar",
  color = "Concordance"
  ) +
  ylim(-2,2.5)+
  theme_classic(base_size = 13)
ggsave("concordance_cervlum.png",width=8,height=6,unit="in",dpi = 300)


###########################################
############ thoracic lumbar ############
###########################################
# Keep overlapping genes and relevant columns
thorlum_df <- thorlum %>%
  dplyr::select(geneid, log_fc_main = log_fc, case_main = case) %>%
  inner_join(
    sens_thorlum %>%
      dplyr::select(geneid, log_fc_sens = log_fc, case_sens = case),
    by = "geneid"
  ) %>%
  mutate(
    concordance = case_when(
      case_sens %in% c("A", "B") &
        case_main %in% c("A", "B")        ~ "Upregulated",
      case_sens %in% c("C", "D") &
        case_main %in% c("C", "D")        ~ "Downregulated",
      case_sens == "E" | case_main == "E" ~ "NS",
      TRUE ~ "Discordant"
    )
  )

# Pearson correlation test
cor_result <- cor.test(
  thorlum_df$log_fc_sens,
  thorlum_df$log_fc_main,
  method = "pearson"
)

# Extract statistics
cor_stats <- data.frame(
  n = nrow(thorlum_df),
  correlation = unname(cor_result$estimate),
  t_value = unname(cor_result$statistic),
  df = unname(cor_result$parameter),
  p_value = cor_result$p.value
)
cor_stats

# plot and color by concordance 
annotation <- sprintf(
  "Pearson r = %.3f",
  cor_stats$correlation
)

ggplot(
  thorlum_df,
  aes(
    x = log_fc_main,
    y = log_fc_sens,
    color = concordance
  )
) +
  geom_hline(yintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_point(
    data = thorlum_df %>%
      filter(!concordance %in% c("Upregulated", "Downregulated")),
    color = "grey70",
    size = 2,
    alpha = 0.6
  ) +
  geom_point(
    data = thorlum_df %>%
      filter(concordance %in% c("Upregulated", "Downregulated")),
    aes(color = concordance),
    size = 2,
    alpha = 0.9
  ) +
  annotate(
    "text",
    x = -Inf, y = Inf,
    label = annotation,
    hjust = -0.05, vjust = 1.1,
    size = 4
  ) +
  scale_color_manual(
    values = c(
      "Upregulated"   = "#D73027",
      "Downregulated"   = "#2166AC",
      "Discordant" ="darkgreen",
      "NS" = "grey65"),
  breaks = c("Upregulated", "Downregulated","Discordant")
  ) +
  labs(
  x = expression(log[2]~"[fold change, main analysis]"),
  y = expression(log[2]~"[fold change, sensitivity analysis]"),
  subtitle = "Thoracic vs. Lumbar",
  color = "Concordance"
  ) +
  ylim(-2,2.5)+
  theme_classic(base_size = 13)
ggsave("concordance_thorlum.png",width=8,height=6,unit="in",dpi = 300)


```
