---
title: "RNA Deconvolution"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(data.table)
library(dplyr)
library(ggrepel)
library(ggplot2)
library(readr)
library(stringr)
library(tidyr)
library(ggVennDiagram)
library(readxl)
library(tibble)
library(edgeR)
library(matrixStats)
library(biomaRt)
library(glmmTMB)
library(emmeans)
library(ggpubr)
library(ggeffects)
library(svglite)
```

# load GTEx

```{r}
setwd("/Users/cindywang/Desktop/Kramer/P10GTEx_raw")

# load data 
gtex_counts <- read.delim("gene_reads_2017-06-05_v8_brain_spinal_cord_cervical_c-1.gct.gz", skip = 2)
gtex_meta<-read.table("GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt", sep = "\t", header = TRUE)
gtex_v10_meta <- read_tsv(
  "GTEx_Analysis_v10_Annotations_SampleAttributesDS.txt",
  quote = "",            # ignore quotes so they don't swallow tabs
  comment = "",          # don't treat # as comments
  progress = FALSE,
  show_col_types = FALSE)

# remove gene name and empty column 
gtex_counts<- gtex_counts %>%
  dplyr::select(-1, -3) %>%
  dplyr::rename(Ensembl = Name)

#clean metadata  
#1. remove hardy scale 2. convert 1=male 2=female 3. convert age range to mid point 
gtex_meta <- gtex_meta %>%
  dplyr::select(-4) %>%
  dplyr::mutate(
    sex = dplyr::recode(
      as.character(SEX),
      "1" = "Male",
      "2" = "Female"
    ),
    age_mid = dplyr::recode(
      as.character(AGE),
      "20-29" = "25",
      "30-39" = "35",
      "40-49" = "45",
      "50-59" = "55",
      "60-69" = "65",
      "70-79" = "75"
    )
  ) %>%
  dplyr::select(-SEX, -AGE)

#Extract donor IDs from spinal cord sample columns
spinal_sample_ids <- colnames(gtex_counts)[2:ncol(gtex_counts)]
spinal_sample_ids <- gsub("\\.", "-", spinal_sample_ids) # Convert sample IDs from R-safe format back to GTEx format
spinal_donor_ids <- substr(spinal_sample_ids, 1, 10) # Extract donor IDs from sample IDs
spinal_donor_ids <- stringr::str_remove(spinal_donor_ids, "-$") # Remove trailing dash if present

#Filter subject metadata to spinal cord donors
subject_meta <- gtex_meta %>%
  dplyr::filter(SUBJID %in% spinal_donor_ids)
nrow(subject_meta) #159 

#filter everything meta to spinal cord donors 
v10_meta <- gtex_v10_meta %>%
  dplyr::filter(SAMPID %in% spinal_sample_ids)
nrow(v10_meta) #159 

#create a "donor ID" column in everything meta 
v10_meta$SUBJID<- substr(v10_meta$SAMPID, 1, 10)
strip_trailing_dash <- function(x) {
  x <- trimws(x)
  sub("-$", "", x)   
}
v10_meta$SUBJID<-strip_trailing_dash(v10_meta$SUBJID)

#merge 2 metadata  
GTEx_meta<-merge(v10_meta,subject_meta,by="SUBJID",all.x=TRUE)

#check ensembl ID duplicates 
dup_row<-gtex_counts %>%
  group_by(Ensembl) %>% 
  filter(n() > 1)
dup_row #no dup in ensembl ID 

#rename the collection sites 
GTEx_meta <- GTEx_meta %>%
  mutate(SMCENTER = case_when(
    SMCENTER == "B1, A1" ~ "GTEx B1, A1",
    SMCENTER == "C1, A1" ~ "GTEx C1, A1",
    SMCENTER == "C1, B1, A1" ~ "GTEx C1, B1, A1",
    SMCENTER == "D1, A1" ~ "GTEx D1, A1",
    TRUE ~ SMCENTER
  ))

```


# load NYGC 

```{r}
setwd("/Users/cindywang/Desktop/Kramer/P9ALS/Humphrey data")
#import count data
cervical_counts <- read.table("Cervical_Spinal_Cord_gene_counts.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
thoracic_counts <- read.table("Thoracic_Spinal_Cord_gene_counts.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
lumbar_counts <- read.table("Lumbar_Spinal_Cord_gene_counts.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

#import metadata
lumbar_meta<-read.table("Lumbar_Spinal_Cord_metadata.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
cerv_meta<-read.table("Cervical_Spinal_Cord_metadata.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
thoracic_meta<-read.table("Thoracic_Spinal_Cord_metadata.tsv", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

#filter for controls in meta 
lumbar_meta<-lumbar_meta %>%  filter(disease=="Control")
cerv_meta<-cerv_meta %>%  filter(disease=="Control")
thoracic_meta<-thoracic_meta %>%  filter(disease=="Control")

#combine all metadata 
NYGC_meta<-rbind(lumbar_meta,cerv_meta,thoracic_meta)

#rename site in metadata 
NYGC_meta <- NYGC_meta %>%
  mutate(site_id = case_when(
    site_id == "site_1" ~ "NYGC site 1",
    site_id == "site_2" ~ "NYGC site 2",
    site_id == "site_3" ~ "NYGC site 3",
    site_id == "site_4" ~ "NYGC site 4",
    site_id == "site_5" ~ "NYGC site 5",
    site_id == "site_6" ~ "NYGC site 6",
    site_id == "site_7" ~ "NYGC site 7",
    site_id == "site_8" ~ "NYGC site 8",
    TRUE ~ site_id
  ))

#filter for controls in count data, keep ensembl only 
cervical_counts <- cervical_counts[, c(names(cervical_counts)[1], cerv_meta$rna_id)]
thoracic_counts <- thoracic_counts[, c(names(thoracic_counts)[1],thoracic_meta$rna_id)]
lumbar_counts <- lumbar_counts[, c(names(lumbar_counts)[1],lumbar_meta$rna_id)]

#merge count data 
a<-merge(cervical_counts,thoracic_counts,by="ensembl_id")
NYGC_counts<-merge(a,lumbar_counts,by="ensembl_id")
dim(NYGC_counts) # 58884    81

#check ensembl ID duplicates 
dup_row<-NYGC_counts %>%
  group_by(ensembl_id) %>% 
  filter(n() > 1)
dup_row # no dup

```

# load Genentech 

```{r}
setwd("/Users/cindywang/Desktop/Kramer/P13MoreBulk")

HF40651<-read_tsv("GSE255683/GSM8077424_SAM24436421_NXG14799307.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF40651=count)
HF40655<-read_tsv("GSE255683/GSM8077425_SAM24436422_NXG14799324.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF40655=count)
HF43950<-read_tsv("GSE255683/GSM8077426_SAM24436423_NXG14799319.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF43950=count)
HF43959<-read_tsv("GSE255683/GSM8077427_SAM24436424_NXG14799320.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF43959=count)
HF43960<-read_tsv("GSE255683/GSM8077428_SAM24436425_NXG14799312.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF43960=count)
HF46773<-read_tsv("GSE255683/GSM8077429_SAM24436426_NXG14799323.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF46773=count)
HF46775<-read_tsv("GSE255683/GSM8077430_SAM24436427_NXG14799316.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF46775=count)
HF32327<-read_tsv("GSE255683/GSM8077431_SAM24436428_NXG14799321.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF32327=count)
HF39053<-read_tsv("GSE255683/GSM8077432_SAM24436429_NXG14799315.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF39053=count)
HF37902<-read_tsv("GSE255683/GSM8077433_SAM24436430_NXG14799310.counts_gene_exonic.tab.gz") %>%
  dplyr::select(-width,-rpkm) %>%
  dplyr::rename(HF37902=count)

#merge data frames by "name"
df_list<-list(HF40651, HF40655, HF43950, HF43959, HF43960, HF46773, HF46775, HF32327, HF39053, HF37902)

df_list <- lapply(df_list, function(df) {
  df$name <- as.character(df$name)
  df
})

Genentech_count <- Reduce(function(x, y) left_join(x, y, by = "name"), df_list)
#dim 58302 11 

#metadata 
setwd("/Users/cindywang/Desktop/Kramer/P13MoreBulk")
genentech_meta<-read_excel("GSE255683/meta.xlsx")
genentech_meta<-genentech_meta[c(11:20),c(1,3,4)]

#rewrite M/F into male/female 
genentech_meta <- genentech_meta %>%
  mutate(Sex = case_when(
    Sex == "F" ~ "Female",
    Sex == "M" ~ "Male",
    TRUE ~ Sex
  ))

#remove the dash in the specimen names 
genentech_meta <- genentech_meta %>%
  mutate(Specimen = str_remove_all(Specimen, "-"))

```

#combine meta 

```{r}
GTEx_partialmeta<-GTEx_meta[,c("SAMPID","SUBJID","SMCENTER","SMRIN","age_mid","sex")]
NYGC_partialmeta<-NYGC_meta[,c("rna_id","dna_id","site_id","tissue","age_rounded","sex","rin")]
Genentech_partialmeta<-genentech_meta[,c("Specimen","Sex","Age at Collection")]

#create missing columns
GTEx_partialmeta$level<-rep("Cervical_Spinal_Cord",159) #create level column for GTEx
Genentech_partialmeta$level<-rep("Lumbar_Spinal_Cord",10) #create level column for Genentech
Genentech_partialmeta$donor<-genentech_meta$Specimen #create donor column for Genentech
Genentech_partialmeta$site<-rep("Genentech",10) #create collection site 
Genentech_partialmeta$RIN <- NA #create RIN 
GTEx_partialmeta$source<-rep("GTEx",159) 
NYGC_partialmeta$source<-rep("NYGC",80) 
Genentech_partialmeta$source<-rep("Genentech",10) 

#reorder
GTEx_partialmeta <- GTEx_partialmeta[, c("SAMPID","SUBJID","SMCENTER","SMRIN","level","age_mid","sex","source")]
NYGC_partialmeta<-NYGC_partialmeta[,c("rna_id","dna_id","site_id","rin","tissue","age_rounded","sex","source")]
Genentech_partialmeta<-Genentech_partialmeta[,c("Specimen","donor","site","RIN","level","Age at Collection","Sex","source")]
    
#rename 
colnames(GTEx_partialmeta)<-c("sample","donor","site","RIN","level","age","sex","source")  
colnames(NYGC_partialmeta)<-c("sample","donor","site","RIN","level","age","sex","source")   
colnames(Genentech_partialmeta)<-c("sample","donor","site","RIN","level","age","sex","source")  
                           
#sanity check 
identical(colnames(GTEx_partialmeta),colnames(NYGC_partialmeta))
identical(colnames(GTEx_partialmeta),colnames(Genentech_partialmeta))

#merge 
merge_meta<-rbind(GTEx_partialmeta,NYGC_partialmeta)
merge_meta<-rbind(merge_meta,Genentech_partialmeta)
dim(merge_meta) #249 8

# setwd("/Users/cindywang/Desktop/Kramer/Publication")
# write.csv(merge_meta, "merge_meta.csv", row.names = FALSE)
```

#combine count 

```{r}
dim(gtex_counts) #56200   160
dim(NYGC_counts) #58884    81
dim(Genentech_count) #58302    11

# Rename the columns 
NYGC_counts<-NYGC_counts %>%
  dplyr::rename(Ensembl=ensembl_id)

Genentech_count<-Genentech_count %>%
  dplyr::rename(Ensembl=name)

############ GTEx processing ############
# remove the Y-chromosome PAR (pseudoautosomal region) annotations (n=44)
gtex_counts <- gtex_counts %>%
  filter(!str_ends(Ensembl, "_PAR_Y"))

# remove the ensembl version in GTEx 
gtex_counts <- gtex_counts %>%
  mutate(Ensembl = str_remove(Ensembl, "\\..*$"))

# check dups 
dup_row<-gtex_counts %>%
  group_by(Ensembl) %>% 
  filter(n() > 1)
dup_row # no dup

# Venn 
venndiagram <- ggVennDiagram(
  x = list(
    GTEx       = gtex_counts$Ensembl,
    NYGC       = NYGC_counts$Ensembl,
    Genentech  = Genentech_count$Ensembl
  ))

# ggsave(
#   filename = "venn_diagram.svg",
#   plot = venndiagram,
#   device = svglite::svglite,
#   width = 7,
#   height = 6,
#   units = "in",
#   bg = "white"
# )

# find overlap 
overlap <- intersect(intersect(gtex_counts$Ensembl, NYGC_counts$Ensembl),
                     Genentech_count$Ensembl)

# filter for overlap
gtex_common <- gtex_counts %>%
  filter(Ensembl %in% overlap)

nygc_common <- NYGC_counts %>%
  filter(Ensembl %in% overlap)

genentech_common <- Genentech_count %>%
  filter(Ensembl %in% overlap)

# merge counts 
merge_counts <- gtex_common %>%
  inner_join(nygc_common, by = "Ensembl") %>%
  inner_join(genentech_common, by = "Ensembl") 
# 55740   250

# setwd("/Users/cindywang/Desktop/Kramer/Publication")
# write.csv(merge_counts, "merge_counts_ensembl.csv", row.names = TRUE)
```
# Demographic info 

```{r}
# change level to abbrev 
merge_meta$level <- recode(
  merge_meta$level,
  "Cervical_Spinal_Cord" = "Cervical",
  "Thoracic_Spinal_Cord" = "Thoracic",
  "Lumbar_Spinal_Cord" = "Lumbar"
)

# age distribution by sex 
merge_meta$level <- factor(
  merge_meta$level,
  levels = c("Cervical", "Thoracic", "Lumbar")
)

ggplot(merge_meta, aes(x = age, fill = level)) +
  geom_histogram(
    binwidth = 10,
    boundary = 0,
    position = "dodge",
    color = "white"
  ) +
  scale_fill_manual(
    limits = c("Cervical", "Thoracic", "Lumbar"),
    values = c(
      "Cervical" = "#F28E2B",  # orange
      "Thoracic" = "#FFE699",  # light yellow
      "Lumbar"   = "#4E79A7"   # blue
    )
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  labs(
    x = "Age",
    y = "Count",
    fill = "Level"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 15),
    legend.title = element_text(size = 17),
    legend.text = element_text(size = 15))

ggsave(
  "age x level.svg",
  device = svglite,
  width = 8,
  height = 3
)

# age distribution by level 
ggplot(merge_meta, aes(x = age, fill = sex)) +
  geom_histogram(
    binwidth = 10,
    boundary = 0,
    position = "dodge",
    color = "white"
  ) +
  scale_fill_manual(
    values = c(
      "Male" = "darkgreen",  
      "Female" = "purple"
    )
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  labs(
    x = "Age",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 15),
    legend.title = element_text(size = 17),
    legend.text = element_text(size = 15))

ggsave(
  "age x sex.svg",
  device = svglite,
  width = 8,
  height = 3
)

```

# Filter lowly expressed genes 

```{r}
#load dataset 
setwd("/Users/cindywang/Desktop/Kramer/Publication")
merge_counts<-read.csv("merge_counts_ensembl.csv",header=TRUE,row.names=1)
merge_meta<-read.csv("merge_meta.csv",header=TRUE)

# column to rownames 
a<-merge_counts %>%
  column_to_rownames("Ensembl")

# CPM transformation
cpm <- cpm(a)
cpm_adjusted = as.data.frame(cpm)

#make sure each variable are of the correct type 
all3_meta <- merge_meta %>%
  mutate(
    across(c(source,level,sex,site), as.factor),
    across(c(age),as.numeric),
    across(c(sample, donor), ~ str_trim(as.character(.)))
  )

# drop gene if CPM is missing in 80% or greater of the participants
sampleNumber <- ncol(cpm_adjusted) #249 samples
keep=rowSums(cpm_adjusted >0) >= sampleNumber*0.2
all3_counts = cpm_adjusted[keep,] 
dim(all3_counts)
#33197   249
```


# PCA

```{r}
colnames(all3_counts) <- str_trim(as.character(colnames(all3_counts)))

#align columns of count matrix with metadata samples
colnames(all3_counts)<-gsub("\\.","-",colnames(all3_counts))
all3_meta<-all3_meta[match(colnames(all3_counts),all3_meta$sample),]
all(colnames(all3_counts)==all3_meta$sample) #TRUE

############ PCA pre ############
dge <- DGEList(counts = all3_counts)

# filter lowly expressed genes
keep <- rowSums(cpm(dge) >= 5) >= 3
dge  <- dge[keep, , keep.lib.sizes = FALSE]

# TMM normalization
dge <- calcNormFactors(dge)

# logCPM
logcpm <- cpm(dge, log = TRUE, prior.count = 1)

# select top variable genes
ntop <- min(5000, nrow(logcpm))
top  <- head(order(rowVars(logcpm), decreasing = TRUE), ntop)

# PCA
pc <- prcomp(t(logcpm[top, ]), scale. = TRUE)

scores <- as.data.frame(pc$x) |>
  rownames_to_column("sample") |>
  left_join(all3_meta, by = "sample")

#plot 
ggplot(scores, aes(PC1, PC2, color = source)) +
  geom_point(size = 3, alpha = 0.9) +
  labs(
    title = "PCA (logCPM, TMM-normalized)",
    x = sprintf("PC1 (%.2f%%)", 100 * summary(pc)$importance[2,1]),
    y = sprintf("PC2 (%.2f%%)", 100 * summary(pc)$importance[2,2])
  ) +
  theme_minimal()+
  theme(
      legend.title = element_text(size = 18),
      legend.text  = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18))
# ggsave("3Combined PCA_prebatch_source.png",width=10,height=8,unit="in",dpi = 300)


############ combat-seq ############
#sanity check 
all(colnames(all3_counts)==all3_meta$sample) #true

#create a batch variable 
batch<-factor(all3_meta$source)

#model matrix for biological covars to preserve
covar_mod <- model.matrix(~ level + sex + age, data = all3_meta)

# 4. ComBat-seq
combat_counts <- sva::ComBat_seq(
  as.matrix(all3_counts),
  batch     = batch,
  covar_mod = covar_mod
)


############ PCA post ############
all(colnames(combat_counts)==all3_meta$sample) #TRUE

# Make DGEList from batch-corrected counts
dge_c <- DGEList(counts = combat_counts)

# Filter lowly expressed genes (same rule as before)
keep_c <- rowSums(cpm(dge_c) >= 5) >= 3
dge_c  <- dge_c[keep_c, , keep.lib.sizes = FALSE]

# TMM normalize
dge_c <- calcNormFactors(dge_c)

# logCPM
logcpm_c <- cpm(dge_c, log = TRUE, prior.count = 1)

# Top variable genes
ntop <- min(5000, nrow(logcpm_c))
top  <- head(order(rowVars(logcpm_c), decreasing = TRUE), ntop)

# PCA on samples
pc_c <- prcomp(t(logcpm_c[top, ]), center = TRUE, scale. = TRUE)

# Variance %
pc_var_pct <- round(100 * (pc_c$sdev^2 / sum(pc_c$sdev^2)), 2)

scores_c <- as.data.frame(pc_c$x) %>%
  tibble::rownames_to_column("sample") %>%
  left_join(all3_meta, by = "sample")

ggplot(scores_c, aes(PC1, PC2, color = level)) +
  geom_point(size = 3, alpha = 0.9) +
  labs(
    title = "PCA (After ComBat-seq)",
    x = sprintf("PC1 (%.2f%%)", pc_var_pct[1]),
    y = sprintf("PC2 (%.2f%%)", pc_var_pct[2]),
    color = "source"
  ) +
  theme_minimal()+
  theme(
      legend.title = element_text(size = 18),
      legend.text  = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18))
# ggsave("3Combined PCA_afterbatch_source.png",width=10,height=8,unit="in",dpi = 300)


#combined before and after plots
scores$stage   <- factor("Before ComBat-seq",
                          levels = c("Before ComBat-seq", "After ComBat-seq"))
scores_c$stage <- factor("After ComBat-seq",
                          levels = c("Before ComBat-seq", "After ComBat-seq"))

both <- bind_rows(scores, scores_c)

ggplot(both, aes(PC1, PC2, color = source)) +
  geom_point(size = 3, alpha = 0.9) +
  facet_wrap(~stage,nrow=1) +
  theme_bw()+
  theme(
      strip.text = element_text(size = 24),
      legend.title = element_text(size = 20),
      legend.text  = element_text(size = 20),
      axis.title = element_text(size = 20),
      axis.text = element_text(size=20))
# ggsave("CombinedPCA_source.png",width=16,height=8,unit="in",dpi = 300)
```
# ENSEMBL to gene name 

```{r}
# rownames back to column "ENSEMBL"
combat_counts <- as.data.frame(combat_counts) %>%
  rownames_to_column(var = "ENSEMBL")

# Connect to Ensembl
mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl"
)

# Query Ensembl
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = combat_counts$ENSEMBL,
  mart = mart
)

# Rename columns
colnames(gene_map) <- c("ENSEMBL", "HGNC")
# only 32506 out of 33197 genes have HGNC symbols

# Merge into original dataframe
combat_counts_genename <- combat_counts %>%
  left_join(gene_map, by = "ENSEMBL") 

# Remove empty string ("") or NA
combat_counts_genename <- combat_counts_genename  %>% 
  filter(!is.na(HGNC), HGNC != "")
dim(combat_counts_genename) # 26587   251

```


#CIBERSORTx pre-deconvolution preprocessing 

```{r}
#check duplicated gene names 
dup_row<-combat_counts_genename1 %>%
  group_by(HGNC) %>% 
  filter(n() > 1)
dup_row #1 duplicated gene name: GOLGA8M

#compute average of duplicated genes 
combat_counts_genename1 <- combat_counts_genename %>%
  group_by(HGNC) %>%
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

# sanity check 
anyDuplicated(combat_counts_genename1$HGNC) == 0

#column to row, drop ENSEMBL column 
Deconv_data <- combat_counts_genename1 %>%
  dplyr::select(HGNC, everything()) %>%
  column_to_rownames("HGNC")

#save as txt
write.table(a, file = "Publication_inc_noncoding.txt", sep = "\t", row.names = TRUE, col.names = TRUE)

a <-read.table("Publication_inc_noncoding.txt", header = TRUE, sep = "\t", row.names = 1)

a<-a %>%
  rownames_to_column()
getwd(
)
```

#Data Viz

```{r}
library(ggbreak)
#load data 
all3_meta<-read.csv("merge_meta.csv")
final_prop<-read.csv("CIBERSORTx_Job60_Results.csv") #mid-level, filtered bulk, permutation 100, porportion, ENSEMBL to gene name  
final_abd<-read.csv("CIBERSORTx_Job61_Results.csv") #mid-level, filtered bulk, permutation 100, abundance, ENSEMBL to gene name  

#proportional barplot for each cell type 
#pivot longer 
Deconv_long<-pivot_longer(
  final_prop,
  cols=Motoneurons:Lymphocytes,
  names_to="cell_type",
  values_to="proportion")

Summary <- Deconv_long %>%
  group_by(cell_type) %>%
  summarise(
    average = mean(proportion),
    sd = sd(proportion),
    n = n()
  ) %>%
  mutate(
    average = average * 100,
    sd = sd * 100,
    sem = sd / sqrt(n)
  ) %>%
  arrange(desc(average))

# save plot 
p<-ggplot(Summary, aes(x = reorder(cell_type, average), y = average)) +
  geom_col(fill = "skyblue") +
  geom_errorbar(
    aes(ymin = average - sem, ymax = average + sem),  
    size = 0.8,
    width = 0.3
  ) +
  geom_text(
    aes(label = round(average, 3)),
    hjust = -1,
    size = 5
  ) +
  coord_flip() +
  labs(
    x = "Cell Types",
    y = "Proportion",
    title = "CIBERSORTx Mid-Level Cell Type Proportion"
  ) +
  scale_y_break(c(15, 65), scales = 0.5) +
  scale_y_continuous(
    breaks = c(0, 5, 10, 15, 65, 76, 79),
    labels = c(0, 5, 10, 15, 65, 76, 79)
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 18),
    legend.text  = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text    = element_text(size = 18),
    panel.grid = element_blank()
  )

svglite("Deconv_proportion.svg", width = 10, height = 10, standalone = TRUE, fix_text_size = TRUE)
print(p)
dev.off()


# QC 
setwd("/Users/cindywang/Desktop/Kramer/P13MoreBulk")
#check RMSE distribution 
ggplot(final_prop,aes(x=RMSE))+
  geom_histogram(binwidth=0.005,fill="grey",color="black")+
  labs(title="Distribution of RMSE",
       x="RMSE",
       y="Count")+
  theme_minimal()+
  theme(
      legend.title = element_text(size = 18),
      legend.text  = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18)
    )
# ggsave("RMSE Distribution.png",width=10,height=4,unit="in",dpi = 300)

#check correlation distribution 
ggplot(final_prop,aes(x=Correlation))+
  geom_histogram(binwidth=0.005,fill="grey",color="black")+
  labs(title="Distribution of Correlation",
       x="Correlation",
       y="Count")+
  theme_minimal()+
  theme(
      legend.title = element_text(size = 18),
      legend.text  = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18)
    )
# ggsave("Correlation Distribution.png",width=10,height=4,unit="in",dpi = 300)

# check p value distribution 
ggplot(final_prop,aes(x=P.value))+
  geom_histogram(binwidth=0.0001,fill="grey",color="black")+
  labs(title="Distribution of P values",
       x="P Value",
       y="Count")+
  theme_minimal()+
  theme(
      legend.title = element_text(size = 18),
      legend.text  = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18)
    )
# ggsave("P value Distribution.png",width=10,height=4,unit="in",dpi = 300)

```

#Data cleanup 

```{r}
#change the dots into dashes in the sample names
final_abd$Mixture <- gsub("\\.", "-", final_abd$Mixture)

#reorder the 2 donor id columns to match
final_abd<-final_abd[match(all3_meta$sample,final_abd$Mixture),]
all(final_abd$Mixture==all3_meta$sample) #TRUE

# keep cells with mean proportions >=1%
celltypes <- final_prop %>%
  dplyr::select(-Mixture) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "cell_type", values_to = "mean_prop") %>%
  filter(mean_prop >= 0.01) %>%
  pull(cell_type)
celltypes<-celltypes[1:6]
#"Motoneurons"     "Astrocyte"      "Macrophages"     "Microglia"      "Oligodendrocyte" "Endothelial"    

final_abd_filt<-final_abd%>%
  dplyr::select(Mixture, Oligodendrocyte,Astrocyte,Endothelial,Motoneurons, Microglia, Macrophages)%>%
  dplyr::select(-any_of(c("RMSE","Correlation")))
# 249   7

#combine columns 
data<-left_join(final_abd_filt,all3_meta,by=c("Mixture"="sample"))

#specify the data types 
data<-data%>%
  mutate(
    across(c(sex,level,donor,source),as.factor),
    across(c(age,RIN),as.numeric))
# 249 * 14 
```

# Check Distribution

```{r}
#check distribution of each cell type using facet wrap
par(mfrow=c(2,3))

final_longer<-final_abd_filt %>%
  dplyr::select(-Mixture) %>%
  pivot_longer(
  cols=everything(),
  names_to = "cell_type",
  values_to = "abundance"
)

ggplot(final_longer,aes(x=abundance))+
  geom_histogram(binwidth=0.1,fill="lightblue",color="black")+
  labs(title="Distribution", x="Absolute Abundance",y="Count")+
  facet_wrap(~cell_type,scales="free")+
  theme_classic()+
  theme(
      legend.title = element_text(size = 16),
      legend.text  = element_text(size = 16),
      axis.title = element_text(size = 16),
      strip.text = element_text(size=16)
    )
# setwd("/Users/cindywang/Desktop/Kramer/Publication/Figures")
# ggsave("Cell Type Distribution.png",width=12,height=8,unit="in",dpi = 300)

#Compute the percentage of zeros in each cell type 
zero_percentages <- sapply(final_abd_filt[-1], function(x) {
  sum(x == 0) / length(x) * 100
})
zero_percentages
# Oligodendrocyte       Astrocyte     Endothelial     Motoneurons       Microglia     Macrophages 
#      0.0000000       1.2048193       0.4016064       0.0000000      10.8433735      56.2248996 
```

#Modelling
GAMMA: "Motoneurons"  "Oligodendrocyte" - when no zeros 
TWEEDIE: "Astrocyte"  "Macrophages"  "Microglia"  "Endothelial"  - when zeros exist

```{r}
##### gamma regression #####
#motoneurons 
fit_mot <- glmmTMB(
  Motoneurons ~ sex + age + level + (1 | donor),
  family = Gamma(link = "log"),
  data = data
)

##### tweedie regression #####
#oligo (even tho it doesn't have zero in the data, tweedie has lower AIC)
fit_oligo <- glmmTMB(
  Oligodendrocyte ~ sex + age + level + (1 | donor),
  family = tweedie(link = "log"),
  data = data
)

#astrocyte 
fit_ast <- glmmTMB(
  Astrocyte ~ sex + age + level + (1 | donor),
  family = tweedie(link = "log"),
  data = data
)

#microglia
fit_micro <- glmmTMB(
  Microglia ~ sex + age + level + (1 | donor),
  family = tweedie(link = "log"),
  data = data
)

#endothelial 
fit_end <- glmmTMB(
  Endothelial ~ sex + age + level + (1 | donor),
  family = tweedie(link = "log"),
  data = data
)

#macrophage 
fit_mac <- glmmTMB(
  Macrophages ~ sex + age + level + (1 | donor),
  family = tweedie(link = "log"),
  data = data
)
```

# emmeans 

```{r}
library(broom.mixed)
library(dplyr)
library(purrr)

data$level <- factor(
  data$level,
  levels = c(
    "Cervical_Spinal_Cord",
    "Thoracic_Spinal_Cord",
    "Lumbar_Spinal_Cord"
  )
)

models <- list(
  Astrocytes    = fit_ast,
  Endothelial   = fit_end,
  Oligodendrocytes = fit_oligo,
  Macrophages   = fit_mac,
  Motoneurons   = fit_mot
)

significant_level_comparisons <- imap_dfr(
  models,
  function(fit, cell_type) {
    
    emm <- emmeans(fit, ~ level)
    
    results <- summary(
      contrast(
        emm,
        method = "revpairwise",
        adjust = "tukey"
      ),
      infer = c(TRUE, TRUE),
      type = "response"
    ) |>
      as.data.frame()
    
    results |>
      filter(p.value < 0.05) |>
      transmute(
        Cell_type = cell_type,
        Comparison = contrast,
        Ratio = ratio,
        CI_lower = asymp.LCL,
        CI_upper = asymp.UCL,
        P_value = p.value
      )
  }
)

significant_level_comparisons
```

#marginal means

```{r}
######## level ########
# rewrite levels 
data <- data %>%
  mutate(level = case_when(
    level == "Cervical_Spinal_Cord" ~ "Cervical",
    level == "Thoracic_Spinal_Cord" ~ "Thoracic",
    level == "Lumbar_Spinal_Cord"   ~ "Lumbar",
    TRUE ~ level
  ))
  
# consistent colors + theme
level_cols <- c(
  "Cervical" = "salmon",
  "Thoracic" = "#1E90FF",
  "Lumbar"   = "#2ca02c"
)

theme_pub <- function() {
  theme_bw() +
    theme(
      legend.title = element_text(size = 14,hjust = 0.5),
      legend.text  = element_text(size = 14),
      axis.title   = element_text(size = 14),
      axis.text    = element_text(size = 12)
    )
}

# helper: build sig-bar df for stat_pvalue_manual
make_pairs_df <- function(emm_obj, emm_df) {
  as.data.frame(pairs(emm_obj, adjust = "tukey")) %>%
    mutate(
      group1 = sub(" /.*", "", contrast),
      group2 = sub(".* / ", "", contrast),
      p.signif = case_when(
        p.value <= 0.001 ~ "***",
        p.value <= 0.01  ~ "**",
        p.value <= 0.05  ~ "*",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(p.signif)) %>%                 # only keep sig comparisons
    arrange(p.value) %>%
    mutate(
      y.position = max(emm_df$asymp.UCL, na.rm = TRUE) *
        (1.05 + 0.08 * (row_number() - 1))
    )
}

plot_emm_level <- function(fit, title, ylim = NULL, out = NULL) {
  emm <- emmeans(fit, ~ level, type = "response")
  emm_df <- as.data.frame(emm)
  emm_df$level <- factor(
  emm_df$level,
  levels = c("Cervical", "Thoracic", "Lumbar"))
  pairs_df <- make_pairs_df(emm, emm_df)

  p <- ggplot(emm_df, aes(x = level, y = response, color = level)) +
    geom_point(size = 8) +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), size = 1, width = 0.3) +
    scale_color_manual(values = level_cols,
                       breaks = c("Cervical","Thoracic","Lumbar")) +
    labs(x = "Level", y = "Marginal Mean (Absolute Abundance)", title = title) +
    theme_bw() +
    theme(
      legend.title = element_text(size = 11,hjust=0.5),
      legend.text  = element_text(size = 11),
      axis.title   = element_text(size = 20),
      axis.text    = element_text(size = 18)
    )

  if (nrow(pairs_df) > 0) {
    p <- p + stat_pvalue_manual(
      pairs_df,
      label = "p.signif",
      tip.length = 0.03,
      size = 6
    )
  }

  if (!is.null(ylim)) p <- p + coord_cartesian(ylim = ylim)
  if (!is.null(out)) ggsave(out, p, width = 8, height = 6, units = "in", dpi = 300)
  p
}


# Motoneurons
plot_emm_level(fit_mot, title = "Motoneurons", ylim = c(0, 2),
               out = "Motoneurons_emmeans.png")
# ggsave("Motoneurons_emmeans.svg",device = svglite,width = 5.5,height = 6)

# Oligodrodrocyte 
plot_emm_level(fit_oligo, title = "Oligodendrocyte", 
                     out = "Oligodrndrocyte_emmeans.png")
# ggsave("Oligodrndrocyte_emmeans.svg",device = svglite,width = 5.5,height = 6)

# Endothelial 
plot_emm_level(fit_end, title = "Endothelial", ylim = c(0, 2),
                     out = "Endothelial_emmeans.png")
# ggsave("Endothelial_emmeans.svg",device = svglite,width = 5.5,height = 6)

# Macrophage 
plot_emm_level(fit_mac, title = "Macrophage", ylim = c(0, 2),
                     out = "Macrophage_emmeans.png")
# ggsave("Macrophage_emmeans.svg",device = svglite,width = 5.5,height = 6)


######## age ########
# macrophage 
age_p <- summary(fit_mac)$coefficients$cond["age", "Pr(>|z|)"]
pred_obs <- ggpredict(fit_mac, terms = c("age")) %>% as.data.frame()
ggplot(data, aes(x = age, y = Macrophages)) +
  geom_point(alpha = 0.4) +
  geom_line(data = pred_obs, aes(x = x, y = predicted, color = group),
            linewidth = 1) +
  theme_bw() +
  labs(x = "Age", y = "Absolute Abundance",
       title="Macrophage")+
  ylim(0,0.5)+
  theme(
      plot.title = element_text(size=20,hjust=0.5),
      legend.text = element_text(size = 18), 
      axis.title = element_text(size = 20), 
      axis.text = element_text(size = 18)
  )+ annotate("text", x = Inf, y = Inf, hjust = 1.2, vjust = 1.5,
                            label = paste0("p = ", signif(age_p, 3)), size = 6)
# ggsave("Macrophage_age.svg",device = svglite,width = 7,height = 11)

```
